Towards a Sound and Complete Library for Modal Logics Reasoning in Coq

### 2022-06-20 paper/paper.tex
Finished section on Lobs theorem. Little changes on proof script.

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  The correct specification of requirements is crucial for software development.
  Modal logics are ...
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

@@

In this section we describe a case study of the defined modal logic library.
We proved the L\"ob theorem in the system K4 with modal fixed points.
The importance of L\"ob's theorem is due to its use in provability
logic~\cite{boolos_1994} to abstract details from encodings in G\"odel's
incomleteness results.

@@

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  This work describes an encoding of modal logics using the Coq proof assistant.
  Our formalization differs from previous attempts by providing a uniform
  representation of several systems for modal logics using Coq's rich type
  structure. We illustrate the usefulness of our library in a formalization of
  L\"ob's theorem which closely follows a classical proof of this result.
\end{abstract}

@@

In this section we describe a case study of the defined modal logic library.
We proved the L\"ob theorem in the system K4 with modal fixed points.
The importance of L\"ob's theorem is due to its use in provability
logic~\cite{boolos_1994} to abstract details from encodings in G\"odel's
incompleteness results.

In order to prove L\"ob's theorem proof we need to encode modal
fixed points. Let $\varphi(A)$ be a formula with only one propositional variable
$A$. We define as a modal fixed point of $\varphi$ is a sentence $\Psi$ such that
$\Psi \leftrightarrow \varphi(\Box \Psi)$.
We represent the modal fixed points by definition \mintinline{coq}{fixed_point},
which is parameterized by a system for modal logics, \mintinline{coq}{S}, and a
theory, \mintinline{coq}{g}, and encodes the logical equivalence between a
formula \mintinline{coq}{p} and its fixpoint \mintinline{coq}{f ([] p)}. 
\begin{minted}{coq}
Definition fixed_point S g
   := forall f, exists p,
          (S ; g |-- [! (p <-> f ([] p)) !]).
\end{minted}

Using the definition of fixed points, we can state L\"ob's theorem as follows:
\begin{minted}{coq}
Theorem Lob:
  forall P, fixed_point K4 nil ->
  (K4; nil |-- [! []P -> P !]) ->
  (K4; nil |-- [! P !]).
\end{minted}

The proof starts by assuming an arbitrary formula $P$, the existence of modal fixed
points and that $\Box P \to P$ is provable in K4. These deduction steps
are performed by the following tactic:
\begin{minted}{coq}
intros P FP H1.
\end{minted}

Next, from the existence of modal fixed points for every formula,
we know that there is a formula $\Psi$ such that $\Psi \leftrightarrow (\Box
\Psi \to P)$. The following tactic performs this step of reasoning.
\begin{minted}{coq}
destruct FP with (fun A => [! A -> P !]) as (psi, H2).
\end{minted}
Now, we are able to deduce that $\Psi \to (\Box \Psi\to P)$ from the previous
step result as follows:
\begin{minted}{coq}
assert (K4; nil |-- [! psi -> []psi -> P !]) as H3.
apply modal_ax5 in H2; try repeat constructor ;
   try eassumption.
\end{minted}
The \mintinline{coq}{assert} tactic allows the definition of
auxiliary proof goals. In the previous code piece, we specified
$\vdash \Psi \to (\Box \Psi \to P)$ as a new goal which
generates a new proof hypothesis named \mintinline{coq}{H3}. The
deduction of this fact is immediate from K4 axioms.

From $\vdash \Psi \to (\Box \Psi \to P)$ and the necessitation
rule, we can deduce $\vdash \Box (\Psi \to (\Box \Psi \to P))$
using the following tactics:
\begin{minted}{coq}
assert (K4; nil |-- [! [](psi -> []psi -> P) !]) as H4.
apply Nec; auto.
\end{minted}

Now, using the axiom \textbf{K} we can deduce $\vdash \Box \Psi \to
\Box (\Box \Psi \to P)$. Again, we deduce this fact using the
\mintinline{coq}{assert} tactic.
\begin{minted}{coq}
assert (K4; nil |-- [! []psi -> []([]psi -> P) !]) as H5.
apply modal_axK in H4; try repeat constructor ;
  try eassumption.
\end{minted}

The next proof step involves showing that
$\vdash \Box (\Box \Psi \to P)\to (\Box \Box \Psi \to \Box P)$ holds
by axiom \textbf{K}.
\begin{minted}{coq}
assert (K4; nil |-- [! []([]psi -> P) ->
                       [][]psi -> []P !]) as H6.
eapply Ax with (a := axK ?[X] ?[Y]);
  try repeat constructor.
\end{minted}

Using $\vdash \Box (\Box \Psi \to P)\to (\Box \Box \Psi \to \Box P)$ and
$\vdash \Box \Psi \to\Box (\Box \Psi \to P)$ we can deduce
$\vdash \Box \Psi \to \Box \Box \Psi \to \Box P$ using composition of
the implication.
\begin{minted}{coq}
assert (K4; nil |-- [! []psi -> [][]psi -> []P !]) as H7.
eapply modal_compose; try repeat constructor ;
   try eassumption.
\end{minted}

Now, we can show that $\vdash\Box\Psi\to\Box\Box\Psi$ using
the axioms of K4. The next snippet of tactics perform this
reasoning step.
\begin{minted}{coq}
assert (K4; nil |-- [! []psi -> [][]psi !]) as H8.
apply modal_axK4; try constructor 2.
\end{minted}

From $\vdash \Box\Psi \to \Box P$ and $\vdash \Box P \to P$, we can
conclude $\vdash \Box \Psi \to P$, by composition of implication as
follows.
\begin{minted}{coq}
assert (K4; nil |-- [! []psi -> P !]) as H10.
eapply modal_compose; try repeat constructor ;
   try eassumption.
\end{minted}

Next, we can conclude $\vdash \Box (\Psi \to P) \to \Psi$ from
$\Psi \leftrightarrow (\Box \Psi \to P)$ and the axioms of K4 using
the following tactics:
\begin{minted}{coq}
assert (K4; nil |-- [! ([]psi -> P) -> psi !]) as H11.
apply modal_ax6 in H2; try repeat constructor ;
 try eassumption.
\end{minted}

From $\vdash (\Box\Psi \to P) \to \Psi$ and $\vdash \Box \Psi \to P$
we can conclude $\vdash \Psi$ by modus ponens.
\begin{minted}{coq}
assert (K4; nil |-- psi) as H12.
eapply Mp ; try eassumption.
\end{minted}

By necessitation rule, we can deduce $\vdash \Box \Psi$
\begin{minted}{coq}
assert (K4; nil |-- [! []psi !]) as H13.
apply Nec ; try eassumption.
\end{minted}

Finally, the desired conclusion $\vdash P$ follows from
using the rule modus ponens over $\vdash \Box \Psi \to P$
and $\vdash \Box \Psi$, as presented by the next tactics
snippet.
\begin{minted}{coq}
eapply Mp ; try eassumption.
\end{minted}

This concludes the formalization of the L\"ob's theorem using
the developed library.

@@

The next table summarizes the different systems of modal logics considered by
previous works.
